---
title: "tutorial_v2-02"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_v2-02}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# **COMb Tutorial v2.02**

## **Contents:**
## 1. Setup
### 1.1. Setup environment

## 2. Data Processing and Curation
### 2.1. Data Preprocessing
### 2.2. Postprocessing of Curated Lipidomics Data
### 2.3. Data Normalization
### 2.4. Quality Control

## 3. Statistical Analysis
### 3.1. Violin/Box plots for each compound class per group
#### 3.1.1. Lipidomics
### 3.2. Fold Change
### 3.3. Dimension Reduction
### 3.4. Statistical Tests (Univariate/Multivariate)

## 4. Visualization of Statistical Results
### 4.1. Volcano Plots
### 4.2. Heatmaps


## **Scripts:**
## 1. Analysis Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 500,
  fig.width = 7
)
```
### 1.1. Define Parameters and Upload Data
```{r setup1, echo = TRUE, fig.align= "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Check if libraries are installed
## HiVE
if ("HiVE" %in% unique(installed.packages()[, 1]) == FALSE) {
  devtools::install_github("cschasestevens/HiVE",
    ref = "main",
    build_vignettes = TRUE
  )
}
# Load libraries
library(COMb)
library(HiVE)
library(SummarizedExperiment)
library(dplyr)
# Set working directory
getwd()
setwd("/path/to/project/directory")
```

## 2. Data Processing, Curation, and Normalization
### 2.1. Data Preprocessing
```{r proc1, echo = TRUE, fig.align= "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Preprocess alignment file exported from MS-DIAL
## Compatible with both ionization modes
data_pos <- ms_preproc(
  dat = "name_of_alignment_file",
  ref1 = "ref/sample_list.csv",
  ref2 = "ref/pos_mzrt_list.txt"
)
data_neg <- ms_preproc(
  dat = "name_of_alignment_file",
  ref1 = "ref/sample_list.csv",
  ref2 = "ref/pos_mzrt_list.txt",
  ionmode = "neg"
)
# Save outputs
## Preprocessed RDS
saveRDS(data_pos, "data/02preprocessing/preprocessed_pos.rds")
saveRDS(data_neg, "data/02preprocessing/preprocessed_neg.rds")
## Output feature list for data curation
write.table(
  data_pos[["data_preproc_feat"]],
  "preprocessed_pos_curate.txt",
  sep = "\t",
  row.names = FALSE
)
write.table(
  data_neg[["data_preproc_feat"]],
  "preprocessed_neg_curate.txt",
  sep = "\t",
  row.names = FALSE
)
```

### 2.2. Postprocessing of Curated Lipidomics Data
```{r proc2, echo = TRUE, fig.align= "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Postprocessing of curated lipidomics data
## Combines adducts and filters data based on final feature list
## Converts data into a SummarizedExperiment for normalization,
## dimension reduction, etc.
# Leave metadata column names unchanged if
# subsequently running ms_qc and ms_qc_report!!!
data1 <- ms_postproc(
  dat = "path/to/curated/data_file.csv",
  dat_in = c(
    "data/02preprocessing/preprocessed_pos.rds",
    "data/02preprocessing/preprocessed_neg.rds"
  ),
  add_metadata = list(
    "Study name" = "Study name",
    "Author" = "Your name",
    "Analysis Platform" = "Q-Exactive HFX",
    "Analysis Date" = "2025-08-19",
    "Data processing" = "MS-DIAL 4.9.221218/LCB and COMb v2.00",
    "Feature column name" = "Label",
    "Class column name" = "Class",
    "Group column name" = "Group",
    "Acquisition order column name" = "Order",
    "Sample column name" = "File type",
    "Blank sample name" = "Blank",
    "QC sample name" = "QC",
    "Internal standard prefix" = "iSTD",
    "Batch column name" = "IDMX",
    "Sample ID column name" = "Label"
  )
)
```

### 2.3. Data Normalization
```{r proc3, echo = TRUE, fig.align= "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Load existing SummarizedExperiment from ms_postproc or upload Excel file
# containing data matrix, compound metadata, and sample metadata
## If loading data from Excel sheet:
data1 <- ms_create_exp(
  file_data = "data/data_sheet.xlsx",
  file_md = "data/study_metadata.xlsx",
  md_no_feat = 29,
  md_no_samp = 7
)
colnames(data1) <- colData(data1)[["SampleID"]]
rownames(data1) <- rowData(data1)[["Name"]]
assay(data1, "norm") <- assay(data1, "format")
# Internal standard normalization
## Need to have specific metadata columns
## listed in 2.2. to work successfully
# 1. Split data by ionization mode
# 2. Split data by region
# 3. Normalize data
data1 <- ms_data_norm(data1)
```

### 2.4. Quality Control
```{r proc4, echo = TRUE, fig.align= "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Run QC function
qcdata <- ms_qc(data1)
# Generate report
ms_qc_report(
  data_qc = qcdata,
  study_name = "Study name",
  author = "Author",
  an_plat = "Analysis Platform",
  an_date = "Analysis Date",
  meth_dp = "Data processing",
  meth_nm = "Normalization method",
  outdir = "qc/",
  outname = "lsae_lipidomics"
)
# Remove qc files from tempdir
file.remove(
  list.files(tempdir(), full.names = TRUE)[
    grepl(
      "heat01|dist01|dimr01|sign01|heat02",
      list.files(tempdir(), full.names = TRUE)
    )
  ]
)
# Save data and qcdata as RDS
saveRDS(data1, "data/20260320_data_lipidomics.rds")
## If split into multiple batches
saveRDS(qcdata[["transformed data"]], "data/20260320_data_lipidomics_split.rds") # nolint
```

## 3. Statistical Analysis
### 3.1. Violin/Box plots for each compound class per group
#### 3.1.1. Lipidomics
```{r statsums01, echo = TRUE, fig.align= "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Filter qcs from data and
# Sort by class, saturation, and chain length
data2 <- data1[
  order(
    rowData(data1)[["Class"]],
    rowData(data1)[["Saturation"]],
    rowData(data1)[["Chain"]]
  ),
  !grepl("Blank|Pool|U", colData(data1)[["Group"]])
]
# Remove internal standards from feature list
data2 <- data2[!grepl("iSTD", rowData(data2)[["Label"]]), ]
# Change NAs in saturation and class to "not specified"
rowData(data2)[["Saturation"]][
  is.na(rowData(data2)[["Saturation"]])
] <- "not specified"
rowData(data2)[["Chain"]][
  is.na(rowData(data2)[["Chain"]])
] <- "not specified"
# Add log2 transformed and pareto-scaled assays to object
## Log2
assay(data2, "log2") <- log2(assay(data2, "norm"))
## Pareto
assay(data2, "scaled") <- apply(
  apply(
    assay(data2, "log2"),
    2,
    function(x) x - mean(x)
  ),
  2,
  function(y) round(y / sqrt(sd(y)), digits = 3)
)
#---- Plots for each class ----
## Create list for each class and plot as panel
lapply(
  unique(rowData(data2)[["Class"]]),
  function(i) {
    # filter data
    ld1 <- data2[rowData(data2)[["Class"]] == i, ]
    sub1 <- ceiling(nrow(ld1) / 20)
    if (sub1 == 1) {
      p1 <- ggpubr::ggarrange(
        plotlist = lapply(
          rowData(ld1)[["Label"]],
          function(j) {
            ms_plot_vio(ld1, comp = j, title1 = j, ptype = "box")
          }
        ),
        ncol = ifelse(nrow(ld1) <= 5, nrow(ld1), 5),
        nrow = ifelse(nrow(ld1) <= 5, 1, ceiling(nrow(ld1) / 5)),
        common.legend = TRUE
      )
      ggplot2::ggsave(
        paste("path/to/directory/lipidomics_", i, ".png", sep = ""), # nolint
        p1, width = 30, height = 20, dpi = 300
      )
    }
    if (sub1 > 1) {
      p1 <- setNames(lapply(
        seq.int(1, sub1, 1),
        function(h) {
          p2 <- rowData(ld1)[["Label"]][
            (1 + 20 * (h - 1)):ifelse(
              (20 * (h - 1)) + 20 > length(rowData(ld1)[["Label"]]),
              length(rowData(ld1)[["Label"]]),
              (20 * (h - 1)) + 20
            )
          ]
          p2 <- ggpubr::ggarrange(
            plotlist = lapply(
              p2,
              function(j) {
                ms_plot_vio(ld1, comp = j, title1 = j, ptype = "box")
              }
            ),
            ncol = ifelse(length(p2) <= 5, length(p2), 5),
            nrow = ifelse(length(p2) <= 5, 1, ceiling(length(p2) / 5)),
            common.legend = TRUE
          )
        }
      ), as.character(seq.int(1, sub1, 1)))
      lapply(
        seq.int(1, length(p1), 1),
        function(k) {
          ggplot2::ggsave(
            paste(
              "path/to/directory/lipidomics_", i,
              "_", names(p1)[[k]], ".png", sep = ""
            ),
            p1[[k]], width = 30, height = 20, dpi = 300
          )
        }
      )
    }
  }
)
# Save RDS containing filtered data
saveRDS(data2, "data/20260325_data_lipidomics_filtered.rds")
```

### 3.2. Fold Change
```{r fold, echo = TRUE, fig.align = "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Calculate fold changes for each compound, class, or lipid class ratio
datafc <- list(
  "standard" = ms_stat_fc(
    exp1 = data2, col_grp = c("Group", "Region")
  ),
  "class" = ms_stat_fc(
    exp1 = data2, col_grp = c("Group", "Region"), fc_method = "class"
  ),
  "HiVE" = ms_stat_fc(
    exp1 = data2, col_grp = c("Group", "Region"), fc_method = "HiVE"
  )
)
lapply(datafc, function(x) head(x))
# Save
lapply(
  names(datafc),
  function(x) {
    write.table(
      datafc[[x]],
      paste(
        "analyze/2026_analysis/table_fc_",
        x, ".txt", sep = ""
      ),
      sep = "\t",
      row.names = FALSE
    )
  }
)
```

### 3.3. Dimension Reduction
```{r dimrd, echo = TRUE, fig.align = "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
set.seed(1234)
# plot PCA/UMAP
ms_dim_rd(
  data2,
  asy = "scaled",
  md_var = "Group"
)
```

### 3.4. Statistical Tests (Univariate/Multivariate)
```{r stat1, echo = TRUE, fig.align = "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Run stats for cells and wash
## Univariate analysis: ANOVA
stat1 <- ms_stat_uni(
  data2,
  asy = "scaled",
  mat_type = "scaled",
  fc_method = "standard"
)
## Univariate analysis: wilcox/t-test
stat1 <- ms_stat_uni(
  data2,
  asy = "scaled",
  mat_type = "scaled",
  fc_method = "standard",
  name_group = c("Group1", "Group2")
)
head(stat1)
## Count significant compounds
setNames(aggregate(
  stat1[["p.value"]],
  list(stat1[["Comparison.fc"]]),
  function(y) {
    d1 <- y < 0.05
    d1 <- length(d1[d1 == TRUE])
  }
), c("Comparison", "no.significant"))
## ChemRICH
stat2 <- ms_stat_crich(
  d_stat = stat1,
  d_ref = as.data.frame(rowData(data4[["cells"]])),
  col_pvalue = "p.value"
)
head(stat2)
```

## 4. Visualization of Statistical Results
### 4.1. Volcano Plots
```{r volp, echo = TRUE, fig.align = "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Plots (Volcano and Heatmap)
### All (for comparisons with at least one significantly altered compound)
lapply(
  seq.int(1, length(unique(stat1[["Type"]])), 1),
  function(x) {
    lapply(
      seq.int(1, length(unique(stat1[["Comparison"]])), 1),
      function(y) {
        tryCatch(
          {
            ggplot2::ggsave(
              paste(
                "analyze/volcano/plot_vol_",
                unique(stat1[["Type"]])[[x]],
                "_",
                gsub("-", "_vs_", unique(stat1[["Comparison"]])[[y]]),
                ".png",
                sep = ""
              ),
              ms_plot_volcano(
                d_stat = stat1[
                  stat1[["Type"]] == unique(
                    stat1[["Type"]]
                  )[[x]] &
                    stat1[["Comparison"]] == unique(
                      stat1[["Comparison"]]
                    )[[y]],
                ],
                p_col = "p adj",
                f_lim = 6,
                comp_name = unique(stat1[["Comparison"]])[[y]],
                x_title = paste(
                  "Increased in",
                  strsplit(
                    unique(stat1[["Comparison"]])[[y]], "-"
                  )[[1]][[2]],
                  "<----->",
                  " Increased in",
                  strsplit(
                    unique(stat1[["Comparison"]])[[y]], "-"
                  )[[1]][[1]],
                  sep = " "
                )
              ),
              width = 8,
              height = 8,
              dpi = 400
            )
          },
          error = function(e) {
            print(
              paste(
                "Error generating plot for",
                unique(stat1[["Type"]])[[x]],
                unique(stat1[["Comparison"]])[[y]]
              )
            )
          }
        )
      }
    )
  }
)
```
### 4.2. Heatmaps
```{r heatp, echo = TRUE, fig.align = "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Heatmaps
h_out <- ms_plot_hm(
  stat_data = stat1,
  h_h = 8,
  h_w = 40,
  cl_c = TRUE,
  cl_r = TRUE
)
png(
  "analyze/plot_heat.png",
  width = 24,
  height = 14,
  units = "cm",
  res = 1000
)
print(h_out)
dev.off()
```

### 4.3. ChemRICH/HiVE Network Plots
```{r netp, echo = TRUE, fig.align = "center", results = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
# Plot standard ChemRICH plot
ms_plot_crich(stat2)
# Plot HiVE enrichment networks
## Annotation only
hive_anno(enr_obj = stat2)
## Enrichment
hive_enrch(enr_obj = stat2, col_sat = "Class")
## Lipid ratio comparison
### Add associated.ratio to stats object
stat1 <- ms_stat_uni(
  data2,
  asy = "scaled",
  mat_type = "scaled",
  fc_method = "HiVE",
  name_group = c("Group1", "Group2")
)
head(stat1)
hive_flux(stat_obj = stat1, col_p = "p.value")
```
